An investigation into using scagnostics to find interesting projections of high-dimensional data

### 2018-10-31 scagnostics.Rmd
Replace sphere with spiral in text.

```{r dataPlotsDensity, fig.width=6, fig.height=7, fig.cap="One dimensional density distribution for all parameters and all data sets considered. Apart from statistical fluctuations we see that for the Pipe datasets the parameters $x_5$ and $x_6$ are pushed towards more extreme values by vetoing points away from the circle outline, and for the Sine dataset the parameter $x_6$ follows a very distinct distribution. Note also that the first five paramters of the Sine dataset appear to be drawn from different normal distributions, which is a result of the rescaling, shifting the modes to different values for each direction."}
pL <- list()
i <- 1
pLabels <- c("Spiral 100", "Spiral 1000", "Pipe 100", "Pipe 1000", "Sine 100", "Sine 1000")
for(ds in list(spiral100, spiral1000, pipe100, pipe1000, sin100, sin1000)){
  distData <- melt(ds)
  pC <- ggplot(distData, aes(value, color=variable)) +
    geom_density() +
    ggtitle(pLabels[i])
  pL[[i]] <- pC
  i <- i+1
}
grid.arrange(pL[[1]], pL[[2]], pL[[3]], pL[[4]], pL[[5]], pL[[6]],
             ncol=2)
```

We next compare the various index values for three views: $x_1$ vs $x_2$, $x_5$ vs $x_6$ and $x_1$ vs $x_6$ (capturing differences from the different distribution in $x_6$). For the Sphere distribution by construction we see only random fluctuations between the different views, which can be sizable in particular for small datasets. Considering next the Pipe distribution, we see that skinny, stringy and dcor2d are increased for the special view, but in particular MIC and TIC index are much larger and may be used to detect this type of structure in the dataset. Moreover, minimising the convex index (or maximising 1-convex) should also allow us to detect the special view. Finally the Sine distribution shows that several of the indices are maximised in the special view, in particular also the splines2d index designed for the detection of fictional dependence.

@@

```{r plotV1V2toV3V4, fig.width=6, fig.height=7, fig.cap="Short tour paths of uninformed views, moving from the projection onto V1-V2 to the projection onto V3-V4, for all considered datasets and index functions. Small samples (p=100) are shown in red, large samples (p=1000) in blue."}
fullResMelt <- melt(fullRes, id=c("t", "name", "size"))
ggplot(fullResMelt, aes(x=t, y=value)) +
  geom_line(aes(color=factor(size))) + 
  facet_wrap(variable~name, ncol=3, scales = "free_y", labeller = label_wrap_gen(multi_line=FALSE)) +
  guides(color=FALSE)
```

### Moving into informative projections
We repeat the same exercise, but now we are interested in how the index values change when going from the view in $x_1$ vs $x_2$ to the view in $x_5$ vs $x_6$, which will also allow us to view the evolution between typical low and high values of the index function. Results are shown in Figure \ref{fig:plannedtour}.
We can see a pretty clear and smooth increase in dcor2d and splines2d as well as MIC and TIC index when approaching the special view of the Sine distribution, which should therefore be efficient to optimise. On the other hand while we also see smooth increase in MIC and TIC when approaching the special view of the Pipe distribution, the increase only becomes apparent much closer to the final projection, making this structure more challenging to detect with a guided tour.
Various indices appear less noisy than seen in \ref{fig:plotV1V2toV3V4}, but note that the y-axis scale is different (typically reaching higher index values in this example, whereas smaller fluctuations are more amplified in the previous views).

@@

## Note on calculation time
\label{sec:timer}
Computing time differs for each index, and depends strongly on the number of data points (as well as probably some dependence on distribution type), here we use as an example the 6-d sphere and calculate the computing time for different indices over 100 interpolated grand tour projections as a function of the sample size.
We see that the evaluation of the spline index shows weak dependence on the sample size, making it the least efficient for very small data sets, but one of the fastest when considering large data sets. The situation is similar for calculating the scagnositics measures, which are however faster to compute on datasets of around 10 points. This is because the implementation is first binning the data points, which is fast compared to the evaluation that requires triangulation of the binned points. The distance correlation based measure exhibits strong increase with sample size, we note however that the implementation also allowes for evaluation on binned data points which we have not considered. Finally for the MINE family the computing time increases exponentially as can be expected from the dependence on the sample size, where we see that the improved $MIC_e$ algorithm is somewhat faster than the original implementation, see also @JMLRv1715308.

>>>>

```{r dataPlotsDensity, fig.width=6, fig.height=7, fig.cap="One dimensional density distribution for all parameters and all data sets considered. Apart from statistical fluctuations we see that for the Pipe datasets the parameters $x_5$ and $x_6$ are pushed towards more extreme values by vetoing points away from the circle outline, and for the Sine (and Spiral) dataset the parameter $x_6$ (and $x_5$) follows a very distinct distribution."}
pL <- list()
i <- 1
pLabels <- c("Spiral 100", "Spiral 1000", "Pipe 100", "Pipe 1000", "Sine 100", "Sine 1000")
for(ds in list(spiral100, spiral1000, pipe100, pipe1000, sin100, sin1000)){
  distData <- melt(ds)
  pC <- ggplot(distData, aes(value, color=variable)) +
    geom_density() +
    ggtitle(pLabels[i])
  pL[[i]] <- pC
  i <- i+1
}
grid.arrange(pL[[1]], pL[[2]], pL[[3]], pL[[4]], pL[[5]], pL[[6]],
             ncol=2)
```

We next compare the various index values for three views: $x_1$ vs $x_2$, $x_5$ vs $x_6$ and $x_1$ vs $x_6$ (capturing differences from the different distribution in $x_6$). For the Spiral distribution the scagnostics indices skinny and stringy are large and close to their theoretic maximum for the special view. On the other hand while dcor2d is larger is also increased for the special view, the value is far from the theoretic maximum and larger values may be found from random fluctuations in other views (CHECK!). MIC and TIC are also found to be sensitive to the spiral distribution. Considering next the Pipe distribution, we see that skinny, stringy and dcor2d are increased for the special view, but in particular MIC and TIC index are much larger and may be used to detect this type of structure in the dataset. Moreover, minimising the convex index (or maximising 1-convex) should also allow us to detect the special view. Finally the Sine distribution shows that several of the indices are maximised in the special view, in particular also the splines2d index designed for the detection of fictional dependence.

@@

```{r plotV1V2toV3V4, fig.width=6, fig.height=7, fig.cap="Short tour paths of uninformed views, moving from the projection onto $x_1$-$x_2$ to the projection onto $x_3$-$x_4$, for all considered datasets and index functions. Small samples (n=100) are shown in red, large samples (n=1000) in blue."}
fullResMelt <- melt(fullRes, id=c("t", "name", "size"))
ggplot(fullResMelt, aes(x=t, y=value)) +
  geom_line(aes(color=factor(size))) + 
  facet_wrap(variable~name, ncol=3, scales = "free_y", labeller = label_wrap_gen(multi_line=FALSE)) +
  guides(color=FALSE)
```

### Moving into informative projections
We repeat the same exercise, but now we are interested in how the index values change when going from the view in $x_1$ vs $x_2$ to the view in $x_5$ vs $x_6$, which will also allow us to view the evolution between typical low and high values of the index function. Results are shown in Figure \ref{fig:plannedtour}.
We can see a pretty clear and smooth increase in dcor2d and splines2d as well as MIC and TIC index when approaching the special view of the Sine distribution, which should therefore be efficient to optimise. On the other hand while we also see smooth increase in MIC and TIC when approaching the special view of the Pipe distribution, the increase only becomes apparent much closer to the final projection, making this structure more challenging to detect with a guided tour. Finally the Spiral appears to be the most challenging, having an even smaller squint angle than the Pipe distribution. While "cmass" appears to be appropriate to identify this view we may actually find larger values of this index for random distributions, making it unfit to use in a guided tour. The scagnostics indices skinny (or 1-convex) appear to be sensitive to the structure at a larger angle compared to MINE indices, but will suffer from the large fluctuations in random views when used in a guided tour.
Various indices appear less noisy than seen in \ref{fig:plotV1V2toV3V4}, but note that the y-axis scale is different (typically reaching higher index values in this example, whereas smaller fluctuations are more amplified in the previous views).

@@

## Note on calculation time
\label{sec:timer}
Computing time differs for each index, and depends strongly on the number of data points (as well as probably some dependence on distribution type), here we use as an example samples of points in a 6-d solid sphere (obtained via the geozoo package [@geozoo]) and calculate the computing time for different indices over 100 interpolated grand tour projections as a function of the sample size. \textcolor{red}{Since projection is fast compared to index evaluation, the number of original dimensions is not very important here. On the other hand different distributions may impact the computing time.}
We see that the evaluation of the spline index shows weak dependence on the sample size, making it the least efficient for very small data sets, but one of the fastest when considering large data sets. The situation is similar for calculating the scagnositics measures, which are however faster to compute on datasets of around 10 points. This is because the implementation is first binning the data points, which is fast compared to the evaluation that requires triangulation of the binned points. The distance correlation based measure exhibits strong increase with sample size, we note however that the implementation also allowes for evaluation on binned data points which we have not considered. Finally for the MINE family the computing time increases exponentially as can be expected from the dependence on the sample size, where we see that the improved $MIC_e$ algorithm is somewhat faster than the original implementation, see also @JMLRv1715308.
